---
title: "Analysis of Prosper loan data from Q4 2005 to Q1 2014"
output: html_document
---

Let's load the data and get an idea about its variables.

```{r}
loans <- read.csv("prosperLoanData.csv")
str(loans)
```

There are approximately 100K records with 81 variables. 

That's a lot of variables! Even with the variable definitions it's hard to keep
track of everything that's going on. Let's plot some data to start getting a 
handle on this data.


```{r}
library(dplyr)
library(ggplot2)
library(tidyr)

# The origination quarter is in format "QX YYYY" which is hard to order so let's
# swap the order of the values.
loans.quarter_split <- separate(loans, 
         col = "LoanOriginationQuarter", 
         into = c("LoanOriginationQuarterOfYear", "LoanOriginationYear"))
loans$LoanOriginationQuarterSwapped = factor(paste(
  loans.quarter_split$LoanOriginationYear, loans.quarter_split$LoanOriginationQuarterOfYear))

loans.originations_by_quarter <- loans %>%
  group_by(LoanOriginationQuarterSwapped) %>%
  summarize(NumLoansOriginated = n()) %>%
  arrange(LoanOriginationQuarterSwapped)

ggplot(loans.originations_by_quarter, aes(LoanOriginationQuarterSwapped, NumLoansOriginated, group = "c")) +
  geom_point() + geom_line()
```

The earliest loans are from 2005-2006, when Prosper was founded and went public.
There's a drop after Q2 2008 (approaching nearly 0 in Q2 2009), likely because 
of the recession. 

The loans started picking up again after that and gradually
increased until the number of originated loans per quarter surpassed 5000, after 
which there was a drop in Q4 2012 and Q1 2013 after which loans started picking 
up again at an even greater rate. 

It would be interesting to see if we could later try to understand the drop and
the subsequent increased growth.