---
title: "Analysis of Prosper loan data from Q4 2005 to Q1 2014"
output: html_document
---

Let's load the data and get an idea about its variables.

```{r, echo=FALSE, message=FALSE}
library(lubridate)
library(plyr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)
library(gridExtra)
```

```{r load_data, cache=TRUE, echo=FALSE}
loans <- read.csv("prosperLoanData.csv")
# Parse origination date.
loans$LoanOriginationDate <- ymd_hms(loans$LoanOriginationDate)
loans$LoanOriginationYear <- year(loans$LoanOriginationDate)

# Original dataset doesn't have loan category labels, let's fix that.
# Source: https://www.prosper.com/Downloads/Services/Documentation/ProsperAPI_Objects_Details.html
category_labels = c(
"Not available",
"Debt consolidation",
"Home improvement",
"Business",
"Personal loan",
"Student use",
"Auto",
"Other",
"Baby & Adoption Loans",
"Boat",
"Cosmetic Procedures",
"Engagement Ring Financing",
"Green Loans",
"Household Expenses",
"Large Purchases",
"Medical/Dental",
"Motorcycle",
"RV",
"Taxes",
"Vacation",
"Wedding Loans")
loans$ListingCategory = factor(
  loans$ListingCategory..numeric., labels=category_labels)

# Add a categorical variable for loan amount range.
amount_bucket_width <- 2000
amount_breaks <- seq(0, 35000, amount_bucket_width)
amount_bucket_labels <- sapply(amount_breaks[-1], 
       function(x) {
         sprintf("%s - %s", dollar(x - amount_bucket_width), dollar(x))
       })
loans$LoanOriginalAmountBucket <- cut(
  loans$LoanOriginalAmount, 
  breaks=amount_breaks, 
  labels=amount_bucket_labels, 
  dig.lab=8)

str(loans)
```

There are approximately 100K records with 81 variables. 

## Number loans originated over time

That's a lot of variables! Even with the variable definitions it's hard to keep
track of everything that's going on. Let's plot some data to start getting a 
handle on this data.

### By quarter

```{r loans_over_quarter, cache=TRUE, fig.width=10, fig.height=6, echo=FALSE}
# The origination quarter is in format "QX YYYY" which is hard to order so let's
# swap the order of the values.
loans.quarter_split <- separate(loans, 
         col = "LoanOriginationQuarter", 
         into = c("LoanOriginationQuarterOfYear", "LoanOriginationYear"))
loans$LoanOriginationQuarterSwapped = factor(paste(
  loans.quarter_split$LoanOriginationYear, 
  loans.quarter_split$LoanOriginationQuarterOfYear))

loans.originations_by_quarter <- loans %>%
  group_by(LoanOriginationQuarterSwapped) %>%
  summarize(NumLoansOriginated = n()) %>%
  arrange(LoanOriginationQuarterSwapped)

ggplot(loans.originations_by_quarter, 
       aes(LoanOriginationQuarterSwapped, NumLoansOriginated, group = "c")) +
  geom_point() + geom_line() + 
  theme(axis.text.x=element_text(angle=-90,vjust=0.5))
```

The earliest loans are from 2005-2006, when Prosper was founded and went public.
There's a drop after Q2 2008 (approaching nearly 0 in Q2 2009), likely because 
of the recession. 

The loans started picking up again after that and gradually
increased until the number of originated loans per quarter surpassed 5000, after 
which there was a drop in Q4 2012 and Q1 2013 after which loans started picking 
up again at an even greater rate. 

It would be interesting to see if we could later try to understand the drop and
the subsequent increased growth.

Grouping by month might reveal more details.

### By month

```{r loans_over_month, cache=TRUE, fig.width=10, fig.height=6, echo=FALSE}
# Extract the month from the loan origination date so we can group by it.
loans$LoanOriginationMonth <- floor_date(loans$LoanOriginationDate, "month")

loans.originations_by_month <- loans %>%
  group_by(LoanOriginationMonth) %>%
  summarize(NumLoansOriginated = n()) %>%
  arrange(LoanOriginationMonth)

ggplot(loans.originations_by_month, 
       aes(LoanOriginationMonth, NumLoansOriginated, group = "c")) +
  geom_point() + 
  geom_line() +
  scale_x_datetime(labels = date_format("%b %Y"), 
                   breaks = date_breaks("3 months")) +
  theme(axis.text.x=element_text(angle=-90,vjust=0.5))
```

Grouped by months, the picture doesn't look much different than before. 
We can tell that the trough in 2013 happed in February. We also see a steep drop
off in March 2014 but it's possible that this data set is not complete for March
2014 as it's the last month in the data set.

What happens if we break down the loans by state? 

### By month, broken down by state

```{r loans_by_month_state, cache=TRUE, fig.width=10, fig.height=6, echo=FALSE}
loans.originations_by_month_state <- loans %>%
  group_by(LoanOriginationMonth, BorrowerState) %>%
  summarize(NumLoansOriginated = n()) %>%
  ungroup() %>%
  arrange(LoanOriginationMonth)

ggplot(loans.originations_by_month_state, 
       aes(LoanOriginationMonth, NumLoansOriginated, fill = BorrowerState)) +
  geom_bar(stat = "identity") + 
  scale_x_datetime(labels = date_format("%b %Y"), 
                   breaks = date_breaks("3 months")) +
  theme(axis.text.x=element_text(angle=-90,vjust=0.5))
```

As expected, this graph is a bit hard to read due to the number of states, but 
it appears that most of the loans are concentrated in a small number of states
so we can try looking at only some of the states. Before we do that, we can 
facet this data by state.

### By month, faceted by state

```{r loans_by_month_faceted_state, cache=TRUE, fig.width=10, fig.height=6, echo=FALSE}
ggplot(loans.originations_by_month_state, 
       aes(LoanOriginationMonth, NumLoansOriginated)) +
  facet_wrap(~BorrowerState) +
  geom_point() + 
  geom_line() +
  scale_x_datetime(labels = date_format("%b %Y"), 
                   breaks = date_breaks("3 months")) +
  theme(axis.text.x=element_text(angle=-90,vjust=0.5))
```

The trends in number of originated loans look similar, but we can see that in 
many states number of originated loans is pretty low. There's also a number of
loans with no state for earlier data.

## Amounts

How much money are people borrowing?

```{r loan_amounts, cache=TRUE, fig.width=10, fig.height=6, echo=FALSE}
ggplot(loans, aes(LoanOriginalAmount)) + 
  geom_bar(binwidth=1000) +
  scale_x_continuous(limits=c(0, 40000), 
                     breaks=seq(0, 40000, 5000), 
                     labels=dollar)
```

Loans amounts appear togo  up to about $35,000.  Most of the loans seem to be 
under $10,000. There are two peaks at around $10,000 and $15,000 and two smaller
peaks around $20,000 and $25,000. Given that the peaks fall around very round 
numbers, I would speculate that people borrowing larger amounts would just pick
a round number. Smaller differences would matter more when borrowing smaller
amounts.

## Success rates

Let's start looking at loan success rates. We first can break down the loans by 
status (grouped by origination date).

### Over time, by month

```{r loans_by_month_status, cache=TRUE, fig.width=10, fig.height=6, echo=FALSE}
loans.originations_by_month_status <- loans %>%
  group_by(LoanOriginationMonth, LoanStatus) %>%
  summarize(NumLoansOriginated = n()) %>%
  arrange(LoanOriginationMonth)

ggplot(loans.originations_by_month_status, 
       aes(LoanOriginationMonth, NumLoansOriginated, color = LoanStatus)) +
  geom_point() + 
  geom_line() +
  scale_x_datetime(labels = date_format("%b %Y"), 
                   breaks = date_breaks("3 months")) +
  theme(axis.text.x=element_text(angle=-90,vjust=0.5))
```

I'd like to explore the difference between succesful and unsuccesful loans, so
I'm going to focus on comparing loans that were completed vs those that have 
been charged off.

```{r loans_by_month_status2, cache=TRUE, fig.width=10, fig.height=6, echo=FALSE}
loans.originations_by_month_status <- loans %>%
  filter(LoanStatus %in% c("Completed", "Chargedoff")) %>%
  group_by(LoanOriginationMonth, LoanStatus) %>%
  summarize(NumLoansOriginated = n()) %>%
  arrange(LoanOriginationMonth)

ggplot(loans.originations_by_month_status, 
       aes(LoanOriginationMonth, NumLoansOriginated, color = LoanStatus)) +
  geom_point() + 
  geom_line() +
  scale_x_datetime(labels = date_format("%b %Y"), 
                   breaks = date_breaks("3 months")) +
  theme(axis.text.x=element_text(angle=-90,vjust=0.5))
```

Loans starting from July 2009 have a Prosper "score" assigned to them which
indicates a probability of a loan "going bad" on a scale from 1-11 with 11 being
the best score. How do Prosper scores break down between completed and charged
off loans?

### Over time, by month (only Chargedoff and Completed)

```{r loans_completed_prosper_score, cache=TRUE, fig.width=10, fig.height=6, echo=FALSE}
loans.completed <- loans %>%
  mutate(ProsperScore = as.factor(ProsperScore)) %>%
  filter(LoanStatus %in% c("Completed", "Chargedoff")) %>%
  filter(LoanOriginationMonth > "2009-07-01")

loans.completed_by_month_status_score <- loans.completed %>%
  group_by(LoanOriginationMonth, LoanStatus, ProsperScore) %>%
  summarize(NumLoansOriginated = n()) %>%
  arrange(LoanOriginationMonth)

ggplot(loans.completed_by_month_status_score, 
       aes(LoanOriginationMonth, NumLoansOriginated, fill = ProsperScore)) +
  facet_wrap(~ LoanStatus, ncol = 1) +
  geom_bar(stat = "identity") + 
  scale_x_datetime(labels = date_format("%b %Y"), 
                   breaks = date_breaks("3 months")) +
  theme(axis.text.x=element_text(angle=-90,vjust=0.5))
```

### Over time, by month (only Chargedoff and Completed, proportional)

Let's normalize the counts to make the graph easier to interpret.

```{r loans_completed_by_month_normalized, cache=TRUE, fig.width=10, fig.height=6, echo=FALSE}
loans.completed_by_month_status <- loans.completed %>%
  group_by(LoanOriginationMonth, LoanStatus) %>%
  summarize(NumLoansOriginatedTotal = n()) %>%
  arrange(LoanOriginationMonth)

# Merge counts by [month, status, score] with counts by [month, status] so that
# we can compute the proportional values.
loans.completed_by_month_status_score_merged <- merge(
  loans.completed_by_month_status_score, 
  loans.completed_by_month_status, 
  by.x = c("LoanOriginationMonth", "LoanStatus"))

loans.completed_by_month_status_score_merged <- mutate(
  loans.completed_by_month_status_score_merged, 
  ProportionOriginated = NumLoansOriginated / NumLoansOriginatedTotal)

ggplot(loans.completed_by_month_status_score_merged, 
       aes(LoanOriginationMonth, ProportionOriginated, fill = ProsperScore)) +
   facet_wrap(~ LoanStatus, ncol = 1) +
   geom_bar(stat = "identity") + 
   scale_x_datetime(labels = date_format("%b %Y"), 
                    breaks = date_breaks("3 months")) +
  theme(axis.text.x=element_text(angle=-90,vjust=0.5))
```

It's a bit surprising that the proportion of scores are similar between the two 
sets. Initially the scores skew high, but become more distributed over time 
presumably as the platform opens up (both with more variety of people coming to 
the platform and as more sophisticated/return hungry lenders enter).

## Charge-off rates

Let's look at some loan properties and see how they affect loan outcomes.

```{r, echo=FALSE}
library(reshape2)
loans.completed = mutate(loans.completed, Term = as.factor(Term))

# Returns a bar chart that shows the proportion of chargedoff and completed 
# loans broken down along the given feature of the dataset.
plot_completed_loans_categorical_breakdown <- function(feature) {
  completed_by_feature <- loans.completed %>%
    group_by_(feature) %>%
    summarize(TotalNumLoans = n())
  
  completes_by_feature_and_status <- loans.completed %>%
    group_by_(feature, "LoanStatus") %>%
    summarize(NumLoans = n())
  
  completes_by_feature_and_status_merged <- merge(
    completed_by_feature, completes_by_feature_and_status, by = c(feature))
  completes_by_feature_and_status_merged <- mutate(
    completes_by_feature_and_status_merged, 
    PropLoans = NumLoans / TotalNumLoans)
  
  return(ggplot(completes_by_feature_and_status_merged, 
                aes_string(feature, "PropLoans", fill = "LoanStatus")) +
    geom_bar(stat = "identity"))
}

```

### By term

```{r, fig.width=10, fig.height=6, echo=FALSE, cache=TRUE}
plot_completed_loans_categorical_breakdown("Term")
```

Here we see that loans with longer terms are charged off more often, but it's
difficult to draw any conclusions. Perhaps the nature of loans is different 
between shorter and longer term loans. It would be interesting to explore if 
there any other variables that correlate with loan term.

Let's plot charged-off loans by various features vs loan term. For categorical
variables, we'll plot proportions attributed to different values of that feature
and for continous variables we'll show a scatter plot.

```{r, fig.width=10, fig.height=6, echo=FALSE, cache=TRUE}
# Returns a bar chart that shows proportions of loans by the given categorical 
# feature for charged-off loans of different terms.
plot_chargedoff_loans_term_breakdown_categorical <- function(feature) {
  charged_off_by_term <- loans.completed %>%
    filter(LoanStatus == "Chargedoff") %>%
    group_by(Term) %>%
    summarize(TotalNumLoans = n())
  
  charged_off_by_term_and_feature <- loans.completed %>%
    filter(LoanStatus == "Chargedoff") %>%
    group_by_("Term", feature) %>%
    summarize(NumLoans = n())
  
  charged_off_by_term_and_feature_merged <- merge(
    charged_off_by_term, charged_off_by_term_and_feature, by = c("Term"))
  charged_off_by_term_and_feature_merged <- mutate(
    charged_off_by_term_and_feature_merged, 
    PropLoans = NumLoans / TotalNumLoans)
  
  return(ggplot(charged_off_by_term_and_feature_merged, 
                aes_string("Term", "PropLoans", fill = feature)) +
    geom_bar(stat = "identity") + 
    guides(fill=FALSE) +
    ggtitle(feature) + 
    theme(axis.title.x = element_blank(), 
          axis.title.y = element_blank(),
          plot.title = element_text(size = rel(0.75)),
          plot.margin = unit(c(-0, 0, 0, 0), "cm")))
}

# Returns a scatter plot of the given feature vs loan term.
plot_chargedoff_loans_term_breakdown_continous <- function(feature) {
  charged_off_loans <- loans.completed %>%
    filter(LoanStatus == "Chargedoff")
  
  return(ggplot(charged_off_loans, 
                aes_string("Term", feature)) +
    geom_point(position=position_jitter(), alpha=0.1) + 
    guides(fill=FALSE) +
    ggtitle(feature) + 
    theme(axis.title.x = element_blank(), 
          axis.title.y = element_blank(),
          plot.title = element_text(size = rel(0.75)),
          plot.margin = unit(c(-0, 0, 0, 0), "cm")))
}

# Returns a grid of plots showing charged off loans broken down by given 
# features.
plot_chargedoff_loans_term_breakdown_multi <- function(features) {
  graphs <- lapply(features, function(x) 
    if(is.factor(loans.completed[[x]])) 
      plot_chargedoff_loans_term_breakdown_categorical(x) 
    else 
      plot_chargedoff_loans_term_breakdown_continous(x))
  graphs$ncol = 3 
  return(do.call(grid.arrange, graphs))
}
```

```{r, fig.width=10, fig.height=20, echo=FALSE, warning=FALSE}
loans.completed$CreditScoreRangeLowerBucket = cut(
  loans.completed$CreditScoreRangeLower, breaks=10)

loans.completed$CreditScoreRangeUpperBucket = cut(
  loans.completed$CreditScoreRangeUpper, breaks=10)

plot_chargedoff_loans_term_breakdown_multi(c(
  "ListingCategory", 
  "Occupation", 
  "EmploymentStatus", 
  "IsBorrowerHomeowner", 
  "IncomeVerifiable", 
  "IncomeRange",
  "LoanOriginalAmount",
  "LoanOriginalAmountBucket",
  "CreditScoreRangeLowerBucket",
  "CreditScoreRangeLower",
  "CreditScoreRangeUpperBucket",
  "CreditScoreRangeUpper",
  "BorrowerAPR",
  "CurrentCreditLines",
  "CurrentDelinquencies",
  "RevolvingCreditBalance",
  "BankcardUtilization",
  "TradesNeverDelinquent..percentage.",
  "DebtToIncomeRatio", # Some data missing
  "StatedMonthlyIncome",
  "MonthlyLoanPayment",
  "Recommendations",
  "Investors"))
```

From the scatter plots, we notice that the number of 12-month loans is 
relatively low compared to 36 and 60 months loans. Given the low number of 12 
month loans, focusing on the difference between 36 and 60 month loans is more
meaningful.

The most interesting plot above, in my eyes, is the scatter plot of loan 
amounts. For the 36-month loans, there's a heavy concentration of loans at $5000
and below. One potential explanation for the difference in charge-off rates is
the difference in the distribution of loan amounts. Loans for higher amounts may
be more likely to be charged off.

### By listing category

```{r, fig.width=10, fig.height=6, echo=FALSE, cache=TRUE}
plot_completed_loans_categorical_breakdown("ListingCategory") +
  theme(axis.text.x=element_text(angle=-90,hjust=0))
```

Charge-off rates vary quite a bit by category. Green loans appear to be the 
riskiest.

### By borrower state

```{r, fig.width=10, fig.height=6, echo=FALSE, cache=TRUE}
plot_completed_loans_categorical_breakdown("BorrowerState")
```

State does appear to affect charge-off reates, but me feeling is that it's 
likely to be caused by different demographics and types of loans.

### By occupation

```{r, fig.width=10, fig.height=6, echo=FALSE, cache=TRUE}
plot_completed_loans_categorical_breakdown("Occupation") + 
  theme(axis.text.x=element_text(angle=-90,hjust=0))
```

There's great variety by occupation, but it's not likely to be an effective 
indicator by itself. It would be worth looking at it in combination with other
characteristics.

### By employment status, homeownership, verifiable income

```{r, fig.width=10, fig.height=6, echo=FALSE, cache=TRUE}
plot_completed_loans_categorical_breakdown("EmploymentStatus")
plot_completed_loans_categorical_breakdown("IsBorrowerHomeowner")
plot_completed_loans_categorical_breakdown("IncomeVerifiable")
```

Same story for these characteristics. It's hard to read into any of these 
individually.

### By borrower rate

```{r, fig.width=10, fig.height=6, echo=FALSE, cache=TRUE}
loans.completed$BorrowerRateBucket = cut(
  loans.completed$BorrowerRate, breaks=10)
plot_completed_loans_categorical_breakdown("BorrowerRateBucket")
```

Here we see that loans with higher rates have a higher charge off rate, which is 
expected as riskier loans would have higher rates.

# Final Plots and Summary

```{r, fig.width=10, fig.height=6, echo=FALSE, cache=TRUE}
loans.originations_by_month_state <- loans %>%
  group_by(LoanOriginationMonth, BorrowerState) %>%
  summarize(NumLoansOriginated = n()) %>%
  ungroup() %>%
  arrange(LoanOriginationMonth) %>%
  mutate(BorrowerState=ifelse(
    BorrowerState == "", "Unknown", as.character(BorrowerState)))

loans.originations_by_state <- loans %>%
  group_by(BorrowerState) %>%
  summarize(NumLoansOriginated = n()) %>%
  arrange(desc(NumLoansOriginated)) %>%
  mutate(BorrowerState=ifelse(
    BorrowerState == "", "Unknown", as.character(BorrowerState)))

top_10_states = loans.originations_by_state[1:10,]$BorrowerState

loans.originations_by_month_state <- loans.originations_by_month_state %>%
  mutate(BorrowerState=ifelse(
    BorrowerState %in% top_10_states, as.character(BorrowerState), "Other")) %>%
  group_by(LoanOriginationMonth, BorrowerState) %>%
  summarize(NumLoansOriginated=sum(NumLoansOriginated))

ggplot(loans.originations_by_month_state, 
       aes(LoanOriginationMonth, NumLoansOriginated, fill = BorrowerState)) +
  geom_bar(stat = "identity") + 
  scale_x_datetime(labels = date_format("%b %Y"), 
                   breaks = date_breaks("3 months")) +
  theme(axis.text.x=element_text(angle=-90,vjust=0.5)) +
  ggtitle("Number of loans originated over time broken down by state") +
  ylab("Number loans") + 
  xlab("") +
  guides(fill=guide_legend(title="Borrower State"))
```

From this graph, we can see the trend in the number of loans originated on the 
Prosper platform. In the beginning the platform was growing steadily, but there
was a sharp drop off in 2008, possibly related to the recession. There appears 
to be a gap in data until 2009 (according to TechCrunch, Prosper stopped 
originating new loans in September 2008 due to SEC scrutiny. Wikipedia indicates
they re-launched in July, 2009). 

Starting in 2009, the loans begun picking up again, only to see another decline 
starting end of 2012. I have not been able to identify the cause of this 
dropoff. After reaching a bottom in early 2013, the platform began growing 
again.

On this graph, we can also see the geographical breakdown of Prosper loans over 
time. Initially, there was a disproportionate amount of loans in Virginia
(reason unclear). Over time the loans became more geographically diverse. 

```{r, fig.width=10, fig.height=6, echo=FALSE, cache=TRUE}
loans.completed_by_month_status <- loans.completed %>%
  group_by(LoanOriginationMonth, LoanStatus) %>%
  summarize(NumLoansOriginatedTotal = n()) %>%
  arrange(LoanOriginationMonth)

loans.completed_by_month_status_score_merged <- merge(
  loans.completed_by_month_status_score, 
  loans.completed_by_month_status, 
  by.x = c("LoanOriginationMonth", "LoanStatus"))

loans.completed_by_month_status_score_merged <- mutate(
  loans.completed_by_month_status_score_merged, 
  ProportionOriginated = NumLoansOriginated / NumLoansOriginatedTotal)

ggplot(loans.completed_by_month_status_score_merged, 
       aes(LoanOriginationMonth, ProportionOriginated, fill = ProsperScore)) +
   facet_wrap(~ LoanStatus, ncol = 1) +
   geom_bar(stat = "identity") + 
   scale_x_datetime(labels = date_format("%b %Y"), 
                    breaks = date_breaks("3 months")) +
  theme(axis.text.x=element_text(angle=-90,vjust=0.5)) +
  ggtitle("Distribution of loans by prosper score over time") +
  ylab("Proportion") + 
  xlab("") +
  guides(fill=guide_legend(title="Prosper Score"))
```

Here we see how the risk quality of loans has changed over time. Initially, the 
data is skewed towards higher quality loans but over time the quality of loans 
becomes more diversified. One possible explanation is that the investor pool 
grew larger as Prosper grew with more investors willing to take on riskier loans
in exchange for higher rates. 

Another possible contributing factor is changes to how the Prosper score is 
computed. Perhaps as Prosper matured, they were able to improve their models 
which might lead to changes in how loans are classified.

We also see that the distribution of charged-off loans trends to lower scores 
over time.

```{r, fig.width=10, fig.height=8, echo=FALSE, cache=TRUE}
g_prop_by_term <- plot_completed_loans_categorical_breakdown("Term") +
  ggtitle("Completed loans broken down by term") +
  ylab("Proportion of completed loans") +
  xlab("Term (Months)") +
  guides(fill=guide_legend(title="Loan Outcome"))

g_chargedoff_prop_by_term <- 
  plot_chargedoff_loans_term_breakdown_categorical("LoanOriginalAmountBucket") + 
  ggtitle("Charge-offs broken down by term and loan amount") +
  ylab("Proportion of charged-off loans") +
  xlab("Term (months)") +
  guides(fill=guide_legend(title="Loan Amount"))

grid.arrange(g_prop_by_term, g_chargedoff_prop_by_term, ncol=1)
```

In the top panel, we see that the proportion of charged-off increases as the 
term of the loan increases.

In the bottom panel, we see a breakdown of the charged-off loans by loan amount.
Longer term loans tend to be for higher amounts, which could explain the higher
charge-off rates, but I'm cautious about drawing any conclusions as there are 
likely other factors that influence the loan amount that could explain away the 
higher charge-off rates.

# Reflection

This has been a challenging data set to work with due to the amount of 
variables. My analysis only scratched the surface of this data. Even looking at 
a small subset of the variables required quite a bit of effort. Someone more 
familiar with the subject might have known which variables were worth focusing 
on. 

Another challenge is that many of the variables in this data set are not 
independent making it hard to make any inferences. One has to be very careful 
when trying to interpret this data as the relationships are tricky. If I were to
do further analysis on this data set, I would focus on trying to understand the
relationship between the various variables before trying to understand their 
relationship to loan success.
